Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs

### 2013-05-30 young_pp_discs.tex
added PdB files

\title[Young protoplanetary discs]{Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs}
%%authors arent splitting over 2 lines so rawlings is pushed off the edge
%\author[Tom Douglas, Paola Caselli, Et al.]{Tom Douglas$^{1}$\thanks{E-mail:
%pytd@leeds.ac.uk}, Paola Caselli$^{1}$, Aaron Boley$^{2}$, Richard Durisen$^{3}$, Tom Hartquist$^{1}$, %John Ilee$^{1}$, Jonathon Rawlings$^{4}$ \\
%$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK \\
%$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
%$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN %47405, USA\\
%%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK\\
%}

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{5,1}$,
A. C. Boley$^{2}$,
T. W. Hartquist$^{1}$,
R. H. Durisen$^{3}$, and
J. M. C. Rawlings$^{4}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK,}} \\
{\small{\it$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA,}}\\
{\small{\it$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA,}}\\
{\small{\it$^{4}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\
{\small{\it$^{5}$School of Physics and Astronomy, University of St Andrews, North Haugh, St Andrews, KY16 9SS, UK,}}\\
}}

%{\parbox{\textwidth}{
%$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK \\
%$^{2}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA\\
%$^{3}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN %47405, USA\\
%$^{4}$Department of Physics & Astronomy, University College London, London WC1E 6BT, UK\\
%}}

@@

\date{Mar 2013}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/fig02.eps}
 \caption{Azimuthaly averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation$V_{\rm rot}=\frac{31.5}{\sqrt{R+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is taken from the same hydrodynamics simulation of a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc, described in I2011. Fig.\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The protostar is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}). {\bf The disc model does not include magnetic fields, which may suppress gravitational instability and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficient magnetisation and alignment between the parent cores rotation axis and magnetic field direction \citep{Joos2012,Li2013}. Thus, the observation of spiral structure would provide strong constraints on the dynamical state of discs. \smallskip }

@@

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. {\bf Photochemistry is not taken into account, based on the assumption that in the early  (Class 0, early ClassI) stages of protostellar accretion and evolution considered here, the protostar/disc system is heavily embedded in a thick envelope of gas and dust. The combination of envelope infall onto the disc and powerful outflows (see e.g. \citealt{Machida2013}) may drastically limit the illumination of the upper layers of young discs by the central protostar. Observations are needed to shed light on the importance of photochemistry at these early stages.}

@@

\subsection{Grid construction} \label{subsec:gridding}
{\bf In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidate points are associated with values of the H$_2$ number density and the number density of the molecular species being simulated at the position of the point in the underlying model. These are then compared against the H$_2$ and molecular density of a reference point in order to decide if they will be included in the grid.

\bf Candidate grid points are selected at random in a cylindrical coordinate system that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) to be used as a threshold.} After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

@@

Fig. \ref{continuum} shows the 1\,mm continuum emission as observed with the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where $g$ is the gas/dust mass ratio, $S_\nu$ is the integrated flux, $d$ is the distance to the source, $\kappa_\nu$ is the dust opacity and $B_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. Using the above equation and integrating the flux from y=0.2 to -0.2 and from x=0.4 to -0.4, we obtain a disc mass of 0.1 and 0.005$\,$M$_\odot$, if the dust temperature is fixed at 40 and 400\,K, respectively. This shows that the disc is optically thick to millimetre wavelengths and millimetre observations underestimate the disc mass by at least a factor of 4 (see also \citealt{Forgan2013}). In young, massive discs such as those we are investigating, longer wavelength continuum observations are necessary for determining the disc mass.\smallskip

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/fig15.eps}
 \caption{Simulated spectra for each of the selected transitions at the positions indicated by the coloured dots in fig. \ref{mom0_maps} (the pink, green and cyan curves are taken from the locations marked by the dots of the same colour). In each diagram, the three spectra are taken from three positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm\,$N (plus yellow, minus pink) where N is the distance from the centre: 10 au for OCS, 20 au for H$_2$CO, 28 au for C$^{17}$O, 38 au for HCO$^+$.}
 \label{spectra}
\end{figure*}

@@

In this paper we have presented radiative transfer simulations of a hybrid model comprising a 0.39$\, M_\odot$ self gravitating disc with a radius of 64$\,$au and spiral density waves, surrounded by an envelope that is a contracting 10$\,M_\odot$ BE-sphere. The main results of these simulations are as follows:\newline
$\bullet$ CASA simulations of this model show that at a distance of 100$\,$pc, extraction of kinematic and structural information from the continuum and molecular lines is possible with ALMA band 7.\\
$\bullet$ Our simulations show that molecular features are predominantly seen in absorption against the hot mid-plane of self-gravitating protoplanetary discs, with emission only being seen where voids in the structure of the disc do not provide a bright continuum source to be absorbed.\\
$\bullet$ The quiescent nature of the envelope around such discs only affects lines within $\pm\,$0.5 km$\,$s$^{-1}$ of the LSR velocity.\\
$\bullet$ The lines studied (OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2) taken together allow all regions of the disc to be sampled and the rotation of the disc to be constrained from the inner edge to the outer edge.\\
$\bullet$ Spiral structures in young embedded discs are detectable at a wide range of inclination angles. They can be resolved spatially at angles close to face on and can be inferred from position velocity diagrams at low inclination angles.\\
{\bf $\bullet$ The dust continuum emission at millimetre wavelengths is optically thick and millimetre observations underestimate the disc mass by significant factors ($>$ 4).\smallskip}

@@

\section*{Acknowledgements}
PC and TWH acknowledge support of successive rolling grants awarded by the UK Science and Technology Funding Council.  
TAD and JDI acknowledge studentships from the Science and Technology Facilities Council of the United Kingdom (STFC), JDI also acknowledges funding from the European Union FP7-2011 under grant agreement no 284405. 
TAD would like to thank Nichol Cunningham for her assistance with CASA and Christian Brinch for his radiative transfer code LIME and for guidance on using it.

>>>>

\title[Young protoplanetary discs]{Simulated Observations of Young Gravitationally Unstable Protoplanetary Discs}

\author[T.~A.~Douglas et al.]
{\parbox{\textwidth}{T. A. Douglas$^{1}$\thanks{E-mail: \texttt{pytd@leeds.ac.uk}},
P. Caselli$^{1}$,
J. D. Ilee$^{2,1}$,
A. C. Boley$^{3}$,
T. W. Hartquist$^{1}$,
R. H. Durisen$^{4}$, and
J. M. C. Rawlings$^{5}$\\
\vspace{0.1cm}\\
{\small{\it$^{1}$School of Physics and Astronomy, University of Leeds, Leeds LS2 9JT, UK,}} \\
{\small{\it$^{2}$School of Physics and Astronomy, University of St Andrews, North Haugh, St Andrews, KY16 9SS, UK,}}\\
{\small{\it$^{3}$Department of Astronomy, University of Florida, 211 Bryant Space Center, PO Box 112055, USA,}}\\
{\small{\it$^{4}$Department of Astronomy, Indiana University, 727 East 3rd Street, Swain West 319, Bloomington, IN 47405, USA,}}\\
{\small{\it$^{5}$Department of Physics \& Astronomy, University College London, London WC1E 6BT, UK}}\\
}}

@@

\date{Accepted May 2013}

@@

\begin{figure}
 \includegraphics[width=84mm]{Figures/model/fig02.eps}
 \caption{Azimuthally averaged rotational velocity in the disc mid-plane. The best fit curve follows the equation$V_{\rm rot}=\frac{31.5}{\sqrt{R+1.68}}+0.518$ .}
 \label{velocity}
\end{figure}

@@

The pre-stellar core structure is maintained down to a radius of 80\,au, within which the model of the young protoplanetary disc is used. The disc structure is taken from the same hydrodynamics simulation of a 1$\,$M$_\odot$ protostar surrounded by a 0.39$\,$M$_\odot$ disc, described in I2011. Fig.\ref{velocity} shows the azimuthally averaged rotational velocity in the disc plane, together with the best fit curve, which will be used in Sec.~\ref{sec:alma_predictions} to compare with position velocity diagrams. The system is envisaged to be at a very early stage of evolution and embedded in an envelope.  The still growing protostar will eventually become an A star through accretion of most of the disc's mass. The hydrodynamics simulation is run with a proper equation of state for molecular hydrogen \citep{Boley2007}.  The protostar is allowed to move freely in response to disc torques, and radiative cooling is included \citep{Boley2009}. The self-gravitating disc exhibits prominent spiral structure, with H$_2$ number densities in the disc ranging from 10$^4$-10$^{13}$ cm$^{-3}$ and temperatures from 30-400 K (fig. \ref{rhoT}).  The disc model does not include magnetic fields, which may suppress gravitational instability and spiral wave formation by slowing down or even preventing early disc growth in the case of sufficient magnetisation and alignment between the parent cores rotation axis and magnetic field direction \citep{Joos2012,Li2013}. Thus, the observation of spiral structure would provide strong constraints on the dynamical state of discs. \smallskip

@@

\subsection{Chemical structure} \label{subsec:chemical_structure}
Chemical abundances in the disc were taken from I2011, who followed gas-grain chemical processes during the dynamical evolution of the disc. Photochemistry is not taken into account, based on the assumption that in the early (Class 0, early ClassI) stages of protostellar accretion and evolution considered here, the protostar/disc system is heavily embedded in a thick envelope of gas and dust. The combination of envelope infall onto the disc and powerful outflows (see e.g. \citealt{Machida2013}) may drastically limit the illumination of the upper layers of young discs by the central protostar. Observations are needed to shed light on the importance of photochemistry at these early stages.

@@

\subsection{Grid construction} \label{subsec:gridding}
 In order to construct the grid, candidate points are randomly selected from the volume to be simulated. These candidate points are associated with values of the H$_2$ number density and the number density of the molecular species being simulated at the position of the point in the underlying model. These are then compared against the H$_2$ and molecular density of a reference point in order to decide if they will be included in the grid.

Candidate grid points are selected at random in a cylindrical coordinate system that is linearly spaced in z and $\theta$ and logarithmically spaced in r. For each point to be selected, a random number $\alpha$ is drawn from the semi-open set [0,$\,$1) to be used as a threshold. After selection of random coordinates, the hydrogen density and molecular density at the candidate point (n and m, respectively) are compared against the densities of a reference point on the inner edge of the disc (n$_0$ and m$_0$). If $\alpha<\left( \frac{n}{n_0} \right)^{0.3}$ or $\alpha< \left( \frac{m}{m_0} \right)^{0.3}$ then the point is selected for use. Otherwise another r, $\theta$, z co-ordinate is selected and it becomes the candidate point. 20\% of these selected points are forced to be at radii greater than $\sqrt{R_{min}R_{max}}$ (where $R_{min}$ and $R_{max}$ are the inner and outer radii of the model) in order to stop too many of the selected points clustering in the high density disc and leaving the envelope under sampled. In addition to this method of selection, 5\% of the points are linearly distributed in x, y and z with no bias with regards to density or abundance. This provides a minimum level of sampling for the large low density regions in the outer parts of the simulated volume. See fig. \ref{points} for an example of the points distribution in r, z. The function comparing the candidate point to the reference point and the candidate point distribution were selected empirically to sample all scales while ensuring that the majority of points are located in the inner disc where the density is higher.  \smallskip

@@

Fig. \ref{continuum} shows the 1\,mm continuum emission as observed with the chosen ALMA configuration. The spiral structure of the disc is clearly visible at this inclination and and it becomes clearer at angles closer to face-on. A mass estimate for the disc was calculated using the equation $M=g S_\nu d^2/\kappa_\nu B_\nu(T_d)$ where $g$ is the gas/dust mass ratio, $S_\nu$ is the integrated flux, $d$ is the distance to the source, $\kappa_\nu$ is the dust opacity and $B_\nu(T_d)$ is the black body function for a given dust temperature \citep{Beltran2006}. Using the above equation and integrating the flux from y=0.2 to -0.2 and from x=0.4 to -0.4, we obtain a disc mass of 0.1 and 0.005$\,$M$_\odot$, if the dust temperature is fixed at 40 and 400\,K, respectively. This shows that the disc is optically thick to millimetre wavelengths and millimetre observations underestimate the disc mass significantly, by factors of 4 or more (see also \citealt{Forgan2013}). In young, massive discs such as those we are investigating, longer wavelength continuum observations are necessary for determining the disc mass.\smallskip

@@

\begin{figure*}
 \includegraphics[width=168mm]{Figures/sim/fig15.eps}
 \caption{Simulated spectra for each of the selected transitions at the positions indicated by the coloured dots in fig. \ref{mom0_maps} (the pink, green and yellow curves are taken from the locations marked by the dots of the same colour). In each diagram, the three spectra are taken from three positions on the Y=0 axis, one is at X=0 (green) and the other 2 are from x=$\pm\,$N (plus yellow, minus pink) where N is the distance from the centre: 10 au for OCS, 20 au for H$_2$CO, 28 au for C$^{17}$O, 38 au for HCO$^+$.}
 \label{spectra}
\end{figure*}

@@

In this paper we have presented radiative transfer simulations of a hybrid model comprising a 0.39$\, M_\odot$ self gravitating disc with a radius of 64$\,$au and spiral density waves, surrounded by an envelope that is a contracting 10$\,M_\odot$ BE-sphere. The main results of these simulations are as follows:\newline
$\bullet$ CASA simulations of this model show that at a distance of 100$\,$pc, extraction of kinematic and structural information from the continuum and molecular lines is possible with ALMA band 7.\\
$\bullet$ Our simulations show that molecular features are predominantly seen in absorption against the hot mid-plane of self-gravitating protoplanetary discs, with emission only being seen where voids in the structure of the disc do not provide a bright continuum source to be absorbed.\\
$\bullet$ The quiescent nature of the envelope around such discs only affects lines within $\pm\,$0.5 km$\,$s$^{-1}$ of the LSR velocity.\\
$\bullet$ The lines studied (OCS 26$\rightarrow$25, H$_2$CO 4$_{04}$$\rightarrow$3$_{03}$, C$^{17}$O 3$\rightarrow$2 and HCO$^+$ 3$\rightarrow$2) taken together allow all regions of the disc to be sampled and the rotation of the disc to be constrained from the inner edge to the outer edge.\\
$\bullet$ Spiral structures in young embedded discs are detectable at a wide range of inclination angles. They can be resolved spatially at angles close to face on and can be inferred from position velocity diagrams at low inclination angles.\\
 $\bullet$ The dust continuum emission at millimetre wavelengths is optically thick and millimetre observations significantly underestimate the disc mass, by a factor of 4 or more.\smallskip

@@

\section*{Acknowledgements}
PC and TWH acknowledge support of successive rolling grants awarded by the UK Science and Technology Funding Council.  
TAD and JDI acknowledge studentships from the Science and Technology Facilities Council of the United Kingdom (STFC), JDI also acknowledges funding from the European Union FP7-2011 under grant agreement no 284405. 
TAD would like to thank Nichol Cunningham for her assistance with CASA and Christian Brinch for his radiative transfer code LIME and for guidance on using it.
We also thank the anonymous referee whose comments improved the initial draft of this manuscript.
